A Machine-checked Proof of the Average-case Complexity of Quicksort in Coq

### 2008-08-29 paper/paper.tex
paper: Add list of expec identities.
darcs-hash:20080829024053-bab43-021b3ec5b841c04ed20f759d951344c517e88da9.gz

%format fcmp = "\circ"
%format match = "\mathbf{match}"
%format with = "\mathbf{with}"
%format end = "\mathbf{end}"
%format fun = "\lambda\hspace{-1mm} "
%format nat = "\mathbb{N}"
%format R = "\mathbb{R}"
%format exists = "\exists "
%format log2ceil = "\log_{2}\hspace{-0.5mm}"
%format over = "\mathbf{over}"
%format INR (a) = a
%format /\ = "\wedge "
%format \/ = "\vee"
%format MonoidMonadTrans.M = "\Varid{MMT}"
%format NatAddMonoid = (nat, 0, +)

@@

@@

Like |expec|, |monoid_expec| gives rise to several identities. One identity that we would like to highlight is
\begin{code}
  monoid_expec_plus: forall (m: Monoid) (h: m -> nat) (A B: Set)
    (f: MMT m NeTreeMonad A) (g: A -> MMT m NeTreeMonad B):
    (forall x y `In` f -> monoid_expec h (g (snd x)) = monoid_expec h (g (snd y))),
    monoid_expec h (f >>= g) =
      monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))),
\end{code}
which states that if one transforms |M_ne_tree| using a monoid |m|, then for a monoid homomorphism |h| from |m| to the additive monoid on |nat|, |monoid_expec h| distributes over |bind|, provided that the expected monoid value of the right hand side does not depend on the computed value of the left hand side. Since |id| is a monoid homomorphism, |monoid_expec_plus| applies to |NDP| and |expec cost|. In section \ref{reduction}, we will use it with another monoid and homomorphism.

@@

After rewriting once more, this time using\begin{code}expec_fcmp: forall f g t, expec (f fcmp g) t = expec f (ne_tree.map g t),\end{code}the goal becomes
\begin{code}
  expec length (ne_tree.map fst (qs_U li)) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}\frac{2}{S (j - i)}}"-}.
\end{code}
We now invoke another lemma which bounds a nondeterministically computed list's expected length by the expected number of occurrences of specific values in that list. More specifically, it states that

\parbox{\textwidth}{\begin{code}
  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)):
    (forall x `In` q, expec (count x) t <= fr i) ->
    (forall x `NotIn` q, expec (count x) t = 0) ->
    NoDuplicates q -> expec length t <= {-"\displaystyle{\sum_{\Varid{i} \in \Varid{q}}\Varid{fr}\ \Varid{i}}"-}.
\end{code}}
The last of the three subgoals thus generated states that there are no duplicate values in |IJ|, which trivially holds. The second subgoal generated is\begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using |expec_fcmp| backwards, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in the following property which is useful in its own right, as we will see later.
\begin{code}sound_cmp_expec_0: forall i j li, (i `NotIn` li \/ j `NotIn` li) ->
  monoid_expec (count (i, j)) (U.qs li) = 0,\end{code}
The proof of |sound_cmp_expec_0|, which we won't show here, is by a separate induction matching |qs|'s recursion.

>>>>

%format fcmp = "\circ"
%format match = "\mathbf{match}"
%format with = "\mathbf{with}"
%format end = "\mathbf{end}"
%format fun = "\lambda\hspace{-1mm} "
%format nat = "\mathbb{N}"
%format R = "\mathbb{R}"
%format exists = "\exists "
%format log2ceil = "\log_{2}\hspace{-0.5mm}"
%format over = "\mathbf{over}"
%format INR (a) = a
%format /\ = "\wedge "
%format \/ = "\vee"
%format <-> = "\leftrightarrow "
%format MonoidMonadTrans.M = "\Varid{MMT}"
%format NatAddMonoid = (nat, 0, +)

@@

|expec| gives rise to a host of identities, including

\begin{code}
  0 <= expec f t
  expec (fun x => f x + g x) t = expec f t + expec g t.
  expec ((* c) fcmp f) = (* c) fcmp expec f

  (forall x `In` t -> f x <= g x) -> expec f t <= expec g t.
  (forall x `In` t -> f x = c) -> expec f t = c
  (forall x, x `In` t -> f x = 0) <-> expec f t = 0

  expec f (t >>= (ret fcmp g)) = expec (f fcmp g) t = expec f (ne_tree.map g t)
\end{code}

@@

In addition to all the identities |monoid_expec| inherits from |expec|, it has some of its own. One identity worth highlighting states that if one transforms |M_ne_tree| using a monoid |m|, then for a monoid homomorphism |h| from |m| to the additive monoid on |nat|, |monoid_expec h| distributes over |bind|, provided that the expected monoid value of the right hand side does not depend on the computed value of the left hand side:
\begin{code}
  monoid_expec_plus: forall (m: Monoid) (h: m -> nat) (A B: Set)
    (f: MMT m NeTreeMonad A) (g: A -> MMT m NeTreeMonad B):
    (forall x y `In` f -> monoid_expec h (g (snd x)) = monoid_expec h (g (snd y))),
    monoid_expec h (f >>= g) =
      monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))).
\end{code}
Since |id| is a monoid homomorphism, |monoid_expec_plus| applies to |NDP| and |expec cost|. In section \ref{reduction}, we will use it with another monoid and homomorphism.

@@

After rewriting with the last of the identities listed for |expec| in section \ref{nondetexpec}, the goal becomes
\begin{code}
  expec length (ne_tree.map fst (qs_U li)) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}\frac{2}{S (j - i)}}"-}.
\end{code}
We now invoke another lemma which bounds a nondeterministically computed list's expected length by the expected number of occurrences of specific values in that list. More specifically, it states that

\parbox{\textwidth}{\begin{code}
  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)):
    (forall x `In` q, expec (count x) t <= fr i) ->
    (forall x `NotIn` q, expec (count x) t = 0) ->
    NoDuplicates q -> expec length t <= {-"\displaystyle{\sum_{\Varid{i} \in \Varid{q}}\Varid{fr}\ \Varid{i}}"-}.
\end{code}}
It is trivially the case that |IJ| does not contain duplicate values. The second subgoal generated is\begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using  the last of the identities listed for |expec| in section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in the following property which is useful in its own right, as we will see later.
\begin{code}sound_cmp_expec_0: forall i j li, (i `NotIn` li \/ j `NotIn` li) ->
  monoid_expec (count (i, j)) (U.qs li) = 0,\end{code}
The proof of |sound_cmp_expec_0|, which we won't show here, is by a separate induction matching |qs|'s recursion.
